---
title: "Basic_Data_Wrangling"
---
# Basic_Data_Wrangling {#basic_data_wrangling .title .toc-ignore}

# Basic_Data_Wrangling {#basic_data_wrangling-1 .title .toc-ignore}

#### Andreas Venizelos {#andreas-venizelos .author}

#### 2022-03-30 {#section .date}

## Downloading Data

One of the most important lesson in R is filtering and cleaning up you
data, in order to generate an easy to read and to handle dataset. First
of all we need to import a dataset. There are many sources for publicly
avaliable big data such as: <https://genome.ucsc.edu/> ,
<https://www.ensembl.org/>, <https://www.encodeproject.org/>. I have
already downloaded a dataset for you. You can download the **txt file**
here:

Download txt file

## Install Packages and Loading Libraries

First step is to install packages either from CRAN, Bioconductor or from
Github and to load the libraries.

``` {r}
library(data.table)
library(dplyr)
library(tidyverse)
library(tm)
```

## Importing Data

Now we will go ahead to import our dataset. There are many ways to
import a dataset depending on the file type and the size of the file.
Here I am going to show you two ways of importing a dataset with the
latter one to be preferred for larger datasets.

### Method 1

``` {r}
filename <- "Encode_HMM_data.txt"
my_data <- read.csv(filename, sep=" ", header=FALSE)
```

### Method 2

``` {r}
library(data.table)
my_data <- fread(filename)
```

``` {r}
head(my_data)
```

    ##      V1    V2    V3                V4 V5 V6    V7    V8          V9
    ## 1: chr1 10000 10600 15_Repetitive/CNV  0  . 10000 10600 245,245,245
    ## 2: chr1 10600 11137 13_Heterochrom/lo  0  . 10600 11137 245,245,245
    ## 3: chr1 11137 11737       8_Insulator  0  . 11137 11737  10,190,254
    ## 4: chr1 11737 11937       11_Weak_Txn  0  . 11737 11937 153,255,102
    ## 5: chr1 11937 12137   7_Weak_Enhancer  0  . 11937 12137   255,252,4
    ## 6: chr1 12137 14537       11_Weak_Txn  0  . 12137 14537 153,255,102

``` {r}
library(dplyr)
glimpse(my_data)
```

    ## Rows: 571,339
    ## Columns: 9
    ## $ V1 <chr> "chr1", "chr1", "chr1", "chr1", "chr1", "chr1", "chr1", "chr1", "ch…
    ## $ V2 <int> 10000, 10600, 11137, 11737, 11937, 12137, 14537, 20337, 22137, 2293…
    ## $ V3 <int> 10600, 11137, 11737, 11937, 12137, 14537, 20337, 22137, 22937, 2693…
    ## $ V4 <chr> "15_Repetitive/CNV", "13_Heterochrom/lo", "8_Insulator", "11_Weak_T…
    ## $ V5 <int> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0…
    ## $ V6 <chr> ".", ".", ".", ".", ".", ".", ".", ".", ".", ".", ".", ".", ".", ".…
    ## $ V7 <int> 10000, 10600, 11137, 11737, 11937, 12137, 14537, 20337, 22137, 2293…
    ## $ V8 <int> 10600, 11137, 11737, 11937, 12137, 14537, 20337, 22137, 22937, 2693…
    ## $ V9 <chr> "245,245,245", "245,245,245", "10,190,254", "153,255,102", "255,252…

## Select Columns

We can see that succesfully we have imported our dataset in R studio.
It's a dataframe with *571,339* rows and *9* columns. We will go ahead
and we will select the first *4* columns from our data set and to ignore
the rest of the columns. Again there are many ways of doing that. Here I
can show you *2* ways :

### Method 1

``` {r}
my_data %>% select(V1,V2,V3,V4)
```

    ##           V1        V2        V3                V4
    ##      1: chr1     10000     10600 15_Repetitive/CNV
    ##      2: chr1     10600     11137 13_Heterochrom/lo
    ##      3: chr1     11137     11737       8_Insulator
    ##      4: chr1     11737     11937       11_Weak_Txn
    ##      5: chr1     11937     12137   7_Weak_Enhancer
    ##     ---                                           
    ## 571335: chrX 155251806 155255406 10_Txn_Elongation
    ## 571336: chrX 155255406 155257806       11_Weak_Txn
    ## 571337: chrX 155257806 155258806       8_Insulator
    ## 571338: chrX 155258806 155259606 13_Heterochrom/lo
    ## 571339: chrX 155259606 155260406 15_Repetitive/CNV

### Method 2

``` {r}
my_data[,1:4]
```

    ##           V1        V2        V3                V4
    ##      1: chr1     10000     10600 15_Repetitive/CNV
    ##      2: chr1     10600     11137 13_Heterochrom/lo
    ##      3: chr1     11137     11737       8_Insulator
    ##      4: chr1     11737     11937       11_Weak_Txn
    ##      5: chr1     11937     12137   7_Weak_Enhancer
    ##     ---                                           
    ## 571335: chrX 155251806 155255406 10_Txn_Elongation
    ## 571336: chrX 155255406 155257806       11_Weak_Txn
    ## 571337: chrX 155257806 155258806       8_Insulator
    ## 571338: chrX 155258806 155259606 13_Heterochrom/lo
    ## 571339: chrX 155259606 155260406 15_Repetitive/CNV

``` {r}
sorted_data <- my_data[,1:4]
```

## Adding Column Names

Now we have selected successfully the columns that we need. However, we
need to rename the columns in order to have a more clean data set.

``` {r}
colnames(sorted_data)[1:4] <- c("Chromosome","Start","Stop","Type")
head(sorted_data)
```

    ##    Chromosome Start  Stop              Type
    ## 1:       chr1 10000 10600 15_Repetitive/CNV
    ## 2:       chr1 10600 11137 13_Heterochrom/lo
    ## 3:       chr1 11137 11737       8_Insulator
    ## 4:       chr1 11737 11937       11_Weak_Txn
    ## 5:       chr1 11937 12137   7_Weak_Enhancer
    ## 6:       chr1 12137 14537       11_Weak_Txn

## Final Clean-Up

Now we have a more defined clean data set. However, there is a string
including a number and an underscore as prefix in each character value
of the column *Type*. Below again *2* ways to remove this string:

### Method 1

``` {r}
sorted_data$Type <- gsub('[[:digit:]]+_','', sorted_data$Type)
```

### Method 2

``` {r}
sorted_data$Type <- str_replace_all(sorted_data$Type, "[:digit:]+_","")
head(sorted_data)
```

    ##    Chromosome Start  Stop           Type
    ## 1:       chr1 10000 10600 Repetitive/CNV
    ## 2:       chr1 10600 11137 Heterochrom/lo
    ## 3:       chr1 11137 11737      Insulator
    ## 4:       chr1 11737 11937       Weak_Txn
    ## 5:       chr1 11937 12137  Weak_Enhancer
    ## 6:       chr1 12137 14537       Weak_Txn

And here you have it ! A final clean data set easy to read and work
with.
